---
title: "Public Transportation Accessibility in Twin Cities"
author: "Tina Chen, Shirley Jiang, Cynthia Zhang"
date: "`r format(Sys.Date(),'%e %B, %Y')`"
format: 
  html:
    self-contained: true
editor: visual
---

```{r, echo = FALSE, results = FALSE, message = FALSE, warning=FALSE}
source('cleaning.R')
```

## Introduction

We were motivated to investigate the accessibility of public transportation within the Twin Cities area with the MetroTransit, being heavily influenced by concerns around equality. A large motivation of every community when they decide to invest heavily in public transportation infrastructure is to give lower-income communities the ability to travel around even when they do not own cars of their own, give individuals with disabilities the opportunity to continue their life, so if the public transportation has issues that pose as challenges for accessibility then that devalues the purpose of public transportation.

With our first research question, we researched whether there were more bus stops in locations based on income and population. The assumption was that there would be more stops in areas with higher population densities and areas that were lower income because those are the communities that would theoretically utilize these services the most. We were also interested in narrowing this broad research question about the urban population to looking into racial demographics of the Twin Cities areas to see if the most marginalized communities were fully served. However, this idea changed as there weren’t enough POC groups within the Twin Cities area to create a compelling case. We could then conclude whether public transit was accessible to younger people as they also rely on public transportation because many can’t drive.

The next question we wanted to understand was related to the ridership changes, especially after the COVID-19 pandemic. It is interesting to observe how the ridership has changed after a pandemic where social distancing and quarantining were so important. This is another indirect method of seeing the accessibility of public transportation because you can learn whether or not initiatives for greater accessibility have been ruled back after the pandemic. To answer these questions, in addition to tracking how daily ridership has varied throughout the years, we narrowed it down by delving into how ridership has changed from weekends to weekdays and different transit lines.

Lastly, we wished to look into the job accessibility along the public transit routes. A major use of public transportation would be to serve commuters and community members in all areas of their lives - which jobs are a big part of. If most jobs are located too far from public transit lines then that would defeat a main purpose of the service.

## Data Resources

In this project, we are using multiple data sources： 

- We extracted the data of daily ridership from 2019 to 2023 and the data of all the existing bus stop locations from **Metropolitan Council internal databases** "db_prod_serv_hist" and "ScheduleDB_Current" respectively.

-   The census tract locations and demographic variables, here we are using population, median income, and race, were extracted from the **2022 ACS (American Community Service) 1-year data set** collected by Census Bureau.

-   The information of schools in each census tract was collecting from the **School Program Location data set** from the Minnesota Geospatial Commons website, collected by Minnesota Department of Education.

-   The number of jobs in each census tract is getting from **Workplace Area Characteristics (WAC) year 2021 data file** in Longitudinal Employer-Household Dynamics (LEHD) data structure collected by Census Bureau. In this project, we are using the "Total number of jobs" variable.

For the data cleaning process, we filtered all the data sets to the target regions, Ramesy County and Hennepin County. Then, we joined the bus stop locations to the census data to count the number of bus stop in each tract. All the other data sets share the same identifier GEOID (unique id for each census tract), so we are allowed to join those data together for further development.


## Demographic and Stops Distribution

### Overview

```{r echo=FALSE}
ggplot() +
  geom_sf(data = census2023, fill = "white") +
  geom_sf(data = stops_sf_county, size = 0.1, color = "blue") +
  theme_void() +
  labs(title = "Bus Stops Distribution")
```

```{r echo=FALSE}
New_name <- c("Population","Median Income", "White ", "Black <br> or African american", "American Indian & <br> Alaska Native ", "Asian ", "Native Hawaiian \n Other Pacific Islander ", "School", "Job")

m <- list()
for (I in 1:9) {
  m[[I]] <- mapview(census2023, zcol = name[I],legend = TRUE,layer.name = New_name[I])
}


do.call(leafsync::sync, c(m[c(1:4, 6, 9)], list(ncol = 3)))

```

From the bus stops distribution map above, we can observe that the locations of bus stops mainly cluster at downtown Saint Paul and downtown Minneapolis. And by looking at the demographic factors, we can roughly conclude that the tracts close to downtown have lower median income and more people of color than the tract close to the edge. Although the maps are quite visually appealing, there isn't enough information we can extract from them.

### Regressions

```{r echo=FALSE, warning=FALSE}
cleaned_stops_census <- clean_names(stops_census_join_sum)

model <- lm(number_of_stops ~ population+median_income + white_alone + black_or_african_american_alone+asian_alone + american_indian_and_alaska_native_alone + native_hawaiian_and_other_pacific_islander_alone + total_jobs, cleaned_stops_census)

# par(mfrow = c(2, 2))

cooksD <- cooks.distance(model)
influential <- cooksD[(cooksD > (3 * mean(cooksD, na.rm = TRUE)))]
outlier <- rep(0, cleaned_stops_census%>% nrow() )
outlier[as.numeric(names(influential)) ] <- 1
outlier_info <- cleaned_stops_census %>% cbind(outlier = outlier) %>% filter(outlier == 1)

outlier_info_sf <- st_as_sf(outlier_info)

ggplot()+
  geom_sf(data = census2023, fill = "white")+
  geom_sf(data = outlier_info_sf, fill = "red")+
  geom_sf(data = stops_sf_county, color = "green", size = 0.1, alpha = 0.3) +
  labs(title = "Outliers Distribution") +
  theme_void()

cleaned_stops_census <- cleaned_stops_census %>% filter(!geoid %in% outlier_info$geoid)
```
To construct a robust regression model and better understand the relationship between bus stop distribution and its influential factors, we employed statistical tools to filter out outliers. We use Cook's distance, a metric that measure the impact of individual observations on the regression coefficients. By computing Cook's distance for each observation in the linear regression model, we identified influential data points. Those observations with Cook's distance exceeding three times the mean Cook's distance were flagged as influential, indicating they potentially exert disproportionate influence on the model's outcomes. We consider these influential observations as outliers due to their significant impact. Visualizing these outliers on a map allows us to spatially understand their distribution and potentially uncover patterns or anomalies in the data. 


We analyzed four of the outliers here:

> Tract 1261.02 is located near the U of M campus in Minneapolis which supports the hypothesis that there would be more stops (greater access to public transportation) for individuals who likely use it more - students. It also has the largest population size out of our narrowed-down list. This would also likely be a viable area for professionals to reside away from downtown Minneapolis but still in commuting distance to their jobs.



> Tract 265.14 is located in Plymouth, MN a large suburb near the Twin Cities. That explains their relatively high median income levels. This tract is also mainly where the bus stops are located to get around the suburb, explaining the high level of stops.



> Tract 1044 is located in central Minneapolis which also makes intuitive sense why it is in the top three for number of stops. There is often more accessibility located in the downtown areas of a city to provide ongoers an easy way to move around. There is likely less population because downtowns tend to be where businesses are located, so many workers probably commute to work from other locations.



> Tract 251 is one the largest tract sizes which might explain partly why it has so many bus stops. However, another reason might be because Mall of America is located within this tract which prompts there to be more public transportation around this area to get to the Mall as it is a major destination within the Twin Cities.

```{r echo=FALSE, warning=FALSE}
columns_to_iterate <- c("population", "median_income", "white_alone", "black_or_african_american_alone", "asian_alone", "american_indian_and_alaska_native_alone", "native_hawaiian_and_other_pacific_islander_alone", "number_school", "total_jobs")

correlation_values <- vector("list", length = 9)

for (col_name in columns_to_iterate) {
# calculate correlation values
  cor_value <- cor(cleaned_stops_census[[col_name]], cleaned_stops_census$number_of_stops, use = "complete.obs")
# confidence intervals
  mod_formula_str <- str_c("number_of_stops", "~", col_name)
  mod_form <- as.formula(mod_formula_str)
  mod <- lm(mod_form, data = cleaned_stops_census)
  ci <- confint(mod, level = 0.95)
# table of data
  correlation_values[[col_name]] <- tibble(
    variable = col_name,
    slope = mod$coefficients[variable],
    ci_lower = ci[variable, "2.5 %"],
    ci_upper = ci[variable, "97.5 %"]
  )
}

correlation_df <- bind_rows(correlation_values)
knitr::kable(correlation_df, format = "html")


## zoom in 
correlation_df[-c(6,7,8),] %>% 
  ggplot(aes(x = factor(variable, level = correlation_df$variable), y = slope))+
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2)+
    geom_point(size = 2)+
  labs(x = "Strength of Different Variables in Relation to Number of Stops")+
  theme_minimal()+
  geom_hline( yintercept = 0, color = "red")+
  scale_x_discrete(label = c("population","median \n income", "white \n alone", "black or \n african american", "asian ", "total job"))

# education
correlation_df[8,] %>% 
  ggplot(aes(x = factor(variable, level = correlation_df$variable), y = slope))+
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2)+
    geom_point(size = 2)+
  labs(x = "Strength of Different Variables in Relation to Number of Stops")+
  theme_minimal()+
  geom_hline( yintercept = 0, color = "red")+
  scale_x_discrete(label = c("population","median \n income", "white \n alone", "black or \n african american", "asian ", " total job"))
```

To gain a deeper understanding of how each factor influences the total number of stops per tract, we build 7 linear regression models. From these models, we obtained coefficients for each factor along with their 95% confidence intervals. We then created visualizations to better illustrate these results.

Based on the first plot, we observe a positive correlation between the number of schools and the number of stops. To be more specific, we are 95% confident that adding an additional school in a tract would result in 4-7 more bus stops.

According for the second plot, we find that `Population`, `White alone`, `Black or African American`, and `Total jobs` all have a positive correlation, while `Median income` has a negative correlation. Regarding `Asian`, since its confidence interval crosses 0, we are uncertain about its correlation.

## Pre-Covid and Post-Covid Comparison

To answer our second broad question, we summarized the ridership and number of routes Metro Transit has before, during, and after the pandemic to do the comparison. Here, we treated ridership as people's demand to public transportation and the number of routes as the supply. Noted that those two variables are mutually related.

### Ridership (Demand)

```{r echo=FALSE}
used_ridership_all %>% 
  group_by(nYear) %>%
  mutate(Total_Riders = sum(Total_Riders,na.rm = TRUE)) %>% 
  ggplot(aes(x=nYear, y=Total_Riders))+
  geom_point() +
  geom_line()+
  labs(x = "Year", y = "Total Riders", title = "Ridership Trend Before and After Covid") +
  theme_classic()
```

First, we look at the demand of public transportation. From our visualizations above, by looking at the total ridership across the years, we can see that there was a significant drop in the ridership from year 2019 to year 2021. This is align with the fact that during pandemic, people switch to online working or drive to work to avoid being infected in public. Starting from year 2021, the total ridership keep increasing as people back to working on-site. But, even until now, the ridership haven't recovered to the pre-Covid level.

```{r echo=FALSE}
weekends_weekdays %>% 
  ggplot(aes(x= nYear, y = Total_Riders)) +
  geom_point()+
  geom_line()+
  facet_wrap(~Schedule) +
  labs(x = "Year", y = "Total Riders", title = "Weekday and Weekend Ridership Comparison") +
  theme_classic()
```

If we looking at the ridership in weekdays and weekends respectively, we can observe that during the pandemic, the ridership in weekdays has been greater affected as that in weekends stayed relatively stable. This is consistent with our previous finding that people who take public transportation to work is the group that have been affected the most.

### Route (Supply)

```{r echo=FALSE, message=FALSE, warning=FALSE}
used_ridership_all %>% 
  distinct(nYear,Route) %>% 
  group_by(nYear) %>% 
  mutate(nRoute = n()) %>% 
  ggplot(aes(x=nYear, y= nRoute)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Number of routes", title = "Change in Number of Routes Provided") +
  theme_classic()

used_ridership_all %>%
  group_by(nYear, rte_class) %>%
  summarise(cnt = n_distinct(Route)) %>% 
  filter(rte_class != "Support") %>% 
  ggplot(aes(x = nYear, y = cnt)) +
  geom_point() +
  geom_line() +
  facet_wrap(~rte_class, labeller = labeller(rte_class = c("CommExp" = "Commuter Express", "CommRai" = "Commuter Rail", "CoreLoc" = "Core Local", "SuburbL" = "Suburb Local"))) +
  labs(x = "Year", y = "number of routes", title = "Number of Route Change Across Route Types") +
  theme_classic()
```

Then, we look into the supply side of public transportation. By looking at the total number of routes provided, we observe a sharp fall from year 2020 to year 2021 and the number is keep decreasing until now. This shows that, as the riders choose not to ride the public transportation, the company was also shutting down the routes. It could because the lack of riders and drivers. Even though the pandemic has past, the number of routes hasn't recovered.

Then we dig into each type of routes respectively. Metro Transit has classified all the routes into several types. BRT is the short for bus rapid transit, Commuter Express are the routes for commuting only during weekdays, Commuter Rail is specific for the North Star Line, Core Local are the major routes distributed around the center of cities, LRT is the short for light rail transit which specific for Green Line and Blue Line, and Suburb Local are the routes located at the edge of cities. From the plots above, the number of Commuter Express routes has a sharp drop during the pandemic and the number of Core Local routes also have a significant decrease, while the other types of routes are relatively stable. The visualizations above shows that people who do public transportation commuting and who live close to cities centers have been affected the most.

## Limitations

A large limitation with our research stemmed from the fact that we are drawing on indirect data to make conclusions on equality, which we measured through ridership. Our ridership data didn't include demographic and income data of the riders using the service, so when studying ridership profiles to determine equity, we had to change the research question with data that we did have - namely, data on the average income and demographic makeup of tracts in the Twin Cities area. This way we could indirectly measure the public transit equality through accessibility for marginalized communities based on where the poorest/most diverse tracts were. However, there are always outliers within these tracts who might not be accounted for because we are focusing on the averages for incomes and demographic who might not be serviced as well because they live in a wealthy/less diverse area. This same issue of indirectness was also present with the analysis on the jobs prevalence in the Twin Cities.

Additionally, the lack of data on timing of each ride was another limitation because we had initially wanted to observe effects of public transit delays because the timing might pose a challenge for riders and cause them to steer away from riding public transit which can become a question of accessibility when preventing people with time-sensitive responsibilities from riding. Within that same data set, we were also not able to observe specifically what the ridership levels looked like for stops. This is something that would have come in handy when understanding the impact of certain lines shutting down after COVID and how that might have affected accessibility.

## Conclusion

Our analysis of public transportation in the Twin Cities area reveals important insights into transit usage, ridership changes, and the impact of demographics and economics. We aimed to see how public transit serves diverse communities, especially those facing socioeconomic challenges and disabilities.

We found that bus stops are often located where there are more people and lower incomes, aligning with the goal of serving marginalized communities. However, there are exceptions, like affluent suburbs with many bus stops.

Ridership dropped significantly during the COVID-19 pandemic, especially on weekdays and commuter routes, and hasn't fully recovered. We also noted limitations in our data, such as indirect measures and lack of detail.To improve public transportation equity, policymakers and transit agencies should invest in underserved areas, improve service reliability, and address barriers faced by marginalized groups.

Overall, by refining our understanding of transit accessibility and addressing disparities, we can create a more inclusive transportation system for everyone in the Twin Cities and beyond.